A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-13 paper/paper.tex
make it fit with really small inference rules

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual.
The category contexts $\Xi$ contain object variables with their
categories; the two forms of category context are $\alpha : \cat C$,
containing one variable that can be used both contravariantly and
covariantly, and $alpha : \cat C ; \beta : \cat D$, containing a
contravariant variable $\alpha$ and covariant variable $\beta$.
Finally, we have contexts $\Phi$ of
element variables with their sets, alternating with those sets' object
variables.  A typical $\Phi$ has the shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}), \alpha_{n+1} : C_{n+1}
\]
and represents the composition of the relations $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \mid FIXME \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

Next, we overview our basic judgement forms along with the
\emph{formation conditions} or \emph{presuppositions} (the conditions
under which a judgement ``type checks'').  We have
\begin{itemize}
\item Categories: ${\Gamma\vdash \cat C \isaCat}$ presupposes ${\Gamma \isadtctx}$.

\item Objects/functors: ${\Gamma \pipe \alpha : \cat C \vdash a : \cat D}$ presupposes
  $\Gamma \vdash \cat C\isaCat$ and $\Gamma \vdash \cat D \isaCat$.
Objects are typed with an input object variable
$\alpha : \cat C$ and an output category $\cat D$; in the semantics
these objects are modeled as functors.

\item Sets/profunctors: ${\Gamma \pipe \Xi \vdash S \isaSet}$ presupposes
      $\Gamma \isadtctx$ and $\Gamma \vdash \Xi \boundary$
A set $S$ is typed with respect
to a set context $\Xi$ to describe its covariant/contravariant
dependence on some input objects. These are semantically modeled as
profunctors.

\item Elements/natural transformations: ${\Gamma \pipe \Phi \vdash s : R}$ presupposes
     ${\Gamma \vdash \Phi \isavectx}$
  and $\Gamma\pipe \underline \Phi \vdash R \isaSet$.
  A transformation $s$ has a context of transformation
variable and a single set output. To be well formed, the context and set
must be parameterized by the same contravariant and covariant object
variables. To ensure this, we use a coercion operation $\underline \Phi$
that erases everything in the context but the left-most and right-most
object variables, i.e. $\underline{\alpha : \cat C} = \alpha : \cat C$ and
$\underline{\Phi} = d^-(\Phi);d^+(\Phi)$ otherwise.

\item Meta-language types and terms: $\Gamma \vdash A \isaTy$ presupposes $\Gamma \isadtctx$ and $\Gamma
  \vdash M : A$ presupposes $\Gamma \isadtctx$ and $\Gamma \vdash A
  \isaTy$.
\end{itemize}

@@

\begin{figure}[t]
  \[
  \begin{array}{l}
    \text{Unit:} \quad
    \inferrule{ ~ }{\Gamma \vdash 1 \isaCat}
    \quad
    \inferrule*{ ~ }{\Gamma\pipe\alpha:C \vdash () : 1}\\\\

@@

    \text{Graph of a profuctor:} \\
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash P \isaSet}
    {\Gamma \vdash \graphProf{\alpha}{\beta} P \isaCat}
    \quad
    \inferrule
    {\defaultObCtx \vdash a_- : C_-\and
      \defaultObCtx \vdash a_+ : C_+\and
      \defaultObCtx \vdash s : P[a_-/\alpha;a_+/\beta]}
    {\defaultObCtx \vdash (a_-,a_+,s) : \graphProf{\alpha:\cat C_-}{\beta:\cat C_+} P}
    \\
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha:\cat C_-}{\beta} P}
    {\defaultObCtx \vdash \pi_- a : \cat C_-}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta:C_+} P}
    {\defaultObCtx \vdash \pi_+ a : \cat C_+}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta} P}
    {\defaultObCtx \vdash \pi_e a : P[\pi_-a/\alpha;\pi_+a/\beta]}
    \\\\
    \text{Negative Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat C \isaCat \and \cat C \isSmall}
    {\Gamma \vdash \negPresheaf {\cat C} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash a : \cat C \and
     \Gamma \pipe d^+\Xi \vdash p : \negPresheaf {\cat C}
    }
    {\Gamma \pipe \Xi \vdash \negPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \beta:\cat D \vdash \lambda \alpha:\cat C. R : \negPresheaf {\cat C}}
    \\
    \\
    \text{Positive Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat D \isaCat \and \cat D \isSmall}
    {\Gamma \vdash \posPresheaf {\cat D} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash p : \posPresheaf {\cat D}\and
     \Gamma \pipe d^+\Xi \vdash a : \cat D
    }
    {\Gamma \pipe \Xi \vdash \posPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \alpha:\cat C \vdash \lambda \beta:\cat D. R : \posPresheaf {\cat D}}
  \end{array}
  \]
  \caption{Category Conectives}
  \label{fig:category-connectives}
\end{figure}

@@

Finally, we have the cartesian unit and product sets, which are
analogous to the normal unit and product of types. The most notable
point to emphasize is that in the formation rule for the product, the
two subformulae should have the same covariant and contravariant
dependence. So, as with linear logic, some constructions can
syntactically use a variable more than once and still be ``linear'': the
unit constructor ``uses'' all variables and the product ``uses'' all
variables in both cases.

\begin{figure}
    % unit,
  \[
  \begin{array}{l}

@@

  \\
  \text{Cartesian unit and products:} \quad
      \inferrule{~}{\Gamma\pipe \Xi \vdash 1 \isaSet}
      \quad
      \inferrule{~}{\Gamma\pipe \Phi \vdash () :  1}
      \\ 
    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \\\\ \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}
    \quad
    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}
    \quad
    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}
  \end{array}
  \]
  \caption{Set Connectives}
  \label{fig:set-connectives}
\end{figure}

@@

Finally, we briefly describe the connectives for the ``meta-logic'',
which extends Martin-L\"of type theory with $\Pi$/$\Sigma$ and extensional
identity types (with their standard rules)~\cite{emltt}. We use
extensional identity types so that the description of models is simpler,
but intensional identity types could be used instead.
%
The types we include are what we call \emph{universes} for the object
categorical logic: types of small categories and locally small
categories, functors, profunctors and natural transformations.
%
The rule for the types of small categories and (large) categories are
very similar: any definable category defines an element of type $\Cats$,
and any element of that type can be reflected back into a category. The
only difference for $\smallCats$ is that the categories involved
additionally satisfy $\cat C \isSmall$. Again we elide the $\beta\eta$
principles, which state that $\quoth{-}$ and $\unquoth{-}$ are mutually
inverse.
%
Since every small category $\cat C \isSmall$ is a category $\cat C
\isaCat$, there is a definable inclusion function from $\smallCats$ to
$\Cat$ and the $\beta\eta$ properties ensure that this is a
monomorphism.

@@

\begin{figure}
  \[
  \begin{array}{l}
    \inferrule
    {~}
    {\Gamma\vdash \smallCats}
    \quad
    \inferrule
    {\Gamma \vdash \cat C \isSmall}
    {\Gamma \vdash \quoth {\cat C} : \smallCats}
    \quad
    \inferrule
    {\Gamma \vdash M : \smallCats}
    {\Gamma \vdash \unquoth M \isSmall}
    \\\\
    \inferrule
    {~}
    {\Gamma\vdash \Cats}
    \quad
    \inferrule
    {\Gamma \vdash \cat C \isaCat}
    {\Gamma \vdash \quoth {\cat C} : \Cat}
    \quad
    \inferrule
    {\Gamma \vdash M : \Cat}
    {\Gamma \vdash \unquoth M \isaCat}
    \\ \\

@@

    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash t : P}
    {\Gamma \vdash \pendlambdaXdotY \alpha t : \pendallXdotY \alpha P}
    \quad
    \inferrule
    {\Gamma \vdash M : \pendallXdotY \alpha P\and
     \Gamma \pipe \beta:\cat D \vdash a : \cat C}
    {\Gamma \pipe \beta:\cat D \vdash \pendappXtoY M a : P[a/\alpha]}
  \end{array}
  \]
  \caption{Type Connectives}
\end{figure}

@@

Next, we can prove that two definitions of an adjunction are equivalent:
\begin{lemma} \label{lem:adjunction}
  Let $R : \varr {\cat D}{\cat C}$ and $L : \varr {\cat
    C}{\cat D}$. Then the following data are in bijection:
  \begin{enumerate}
  \item An isomorphism of profunctors $(\punitinXfromYtoZ {\cat D} {L \alpha}{\beta}) \cong (\punitinXfromYtoZ {\cat C} \alpha {R \beta})$
  \item A unit $\eta : \pendallXdotY \alpha {\punitinXfromYtoZ {\cat C} \alpha
    {R(L\alpha)}}$ and co-unit $\varepsilon: \pendallXdotY \beta
    {\punitinXfromYtoZ {\cat D} {L(R(\beta))}{\beta}}$ satisfying triangle
    identities.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Given the forward homomorphism $\text{lr}$, we can construct $\eta =
  \pendlambdaXdotY \alpha \appunary {\text{lr}} \alpha {\punitrefl
    \alpha} {L \alpha}$ and given the unit we can reconstruct the
  forward homomorphism using
  $\text{comp}$ and $\text{fctor}$ as defined in
  Construction\ref{construction:synthetic-composition} as
  $\appbinary {\text{comp}} \alpha
  {\pendappXtoY \eta \alpha} {R(L\alpha)} {(\appunary
    {\text{fctor}(R)}{L \alpha} {f} {\beta})} {R \beta}$.
\end{proof}

Next, we turn to weighted limits, which generalize ordinary limits and
Kan extensions.
\begin{definition}
  For a functor $D : \varr {\cat J}{\cat C}$ and a profunctor $W :
  \harr{\cat K}{\cat J}$, the limit of $D$ weighted by $W$ is (if it
  exists) a functor $\weightedLimitDW D W : \varr{\cat K}{\cat C}$
  with an isomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {(\weightedLimitDW D W) k} \cong \homrallXYtoZ j {W k j} {(\punitinXfromYtoZ {\cat C} \alpha {D j})}$
\end{definition}
This generalizes the usual definition that a morphism into a limit is
a cone over the diagram $(\punitinXfromYtoZ {\cat C} \alpha {D j})$ to
be parameterized by a weight $W k j$.
Then we can prove the famous theorem that right adjoints preserve (weighted) limits:
\begin{theorem}
  If $\weightedLimitDW D W$ exists and is a limit and $R : \varr{\cat
    C}{\cat C'}$ has a left adjoint $L$, then $\lambda
  \kappa. R((\weightedLimitDW D W) \kappa)$ is the limit of $\lambda
  j. R(D j)$ weighted by $W$.
\end{theorem}
\begin{proof}
  \begin{align*}
    \punitinXfromYtoZ {} {\gamma} {R ((\weightedLimitDW D W) \kappa)}
    \cong \punitinXfromYtoZ {}{L \gamma} {(\weightedLimitDW D W) \kappa}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{L \gamma}{D j}}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{\gamma}{R(D j)}}\\
  \end{align*}  
\end{proof}
This is a high level proof in terms of isomorphisms that may be
written in VETT. The first two steps are the instantiation of
assumptions (adjointness, weighted limits). The last step uses the
fact that a natural isomorphisms lift to natural isomorphism of homs
of profunctors. We show the construction of this isomorphism as follows,
giving an example that naturality need not be proved explicitly.
\begin{construction}
  For any $\phi: \homunary \alpha {R' \alpha \beta} \beta {R \alpha
    \beta}$ and $\psi : \homunary \gamma {S \gamma \beta} \beta {S'
    \gamma \beta}$ we can construct a natural transformation $\phi
  \triangleright \psi : \hombinary \gamma {(\homrallXYtoZ \beta {R
      \alpha \beta}{S \gamma \beta})} {\alpha} {R'\alpha\beta} \beta {S'\gamma\beta}$ as
  $\lambdabinary {\gamma} {f} {\alpha} r \beta {\appunary \psi \gamma {(\homrappXtoYatZ f {(\appunary \phi \alpha r \beta)} \beta)} \beta}$. Furthermore if $\phi$ and $\psi$ have inverses, then $\phi^{-1} \triangleright \psi^{-1}$ is the inverse of $\phi \triangleright \psi$.
\end{construction}

@@

\subsection{Extensionality Axioms}

>>>>

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
The category contexts $\Gamma \vdash \Xi \boundary$ contain object
variables with their categories; the two forms of category context are
$\alpha : \cat C$, containing one variable that can be used both
contravariantly and covariantly, and $\alpha : \cat C ; \beta : \cat D$,
containing a contravariant variable $\alpha$ and covariant variable
$\beta$.  Finally, we have contexts $\Phi$ of element variables with
their sets, alternating with those sets' object variables.  A typical
$\Phi$ has the shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}), \alpha_{n+1} : C_{n+1}
\]
and represents the composition of the relations $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \mid \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

Next, we overview our basic judgement forms.  We have
\begin{itemize}
\item Categories: ${\Gamma\vdash \cat C \isaCat}$, where ${\Gamma \isadtctx}$.

\item Objects/functors: ${\Gamma \pipe \alpha : \cat C \vdash a : \cat
  D}$, where $\Gamma \vdash \cat C\isaCat$ and $\Gamma \vdash \cat D
  \isaCat$.  Objects are typed with an input object variable $\alpha :
  \cat C$ and an output category $\cat D$; in the semantics, objects are
  modeled as functors $\cat C \to \cat D$.

\item Sets/profunctors: ${\Gamma \pipe \Xi \vdash S \isaSet}$, where
  $\Gamma \vdash \Xi \boundary$. A set $S$ is typed with respect to a set
  context $\Xi$ to describe its covariant/contravariant dependence on
  some input objects. Sets are semantically modeled as profunctors.

\item Elements/natural transformations: ${\Gamma \pipe \Phi \vdash s :
  R}$, where ${\Gamma \vdash \Phi \isavectx}$ and $\Gamma\pipe
  \underline \Phi \vdash R \isaSet$.  A transformation $s$ has a context
  of transformation variable and a single set output. To be well formed,
  the context and set must be parameterized by the same contravariant
  and covariant object variables. To ensure this, we use a coercion
  operation $\underline \Phi$ that erases everything in the context but
  the left-most and right-most object variables, i.e. $\underline{\alpha
    : \cat C} = \alpha : \cat C$ and $\underline{\Phi} =
  d^-(\Phi);d^+(\Phi)$ otherwise.

\item Meta-language types and terms: $\Gamma \vdash A \isaTy$ and
  $\Gamma \vdash M : A$ as in standard dependent type theory.
\end{itemize}

@@

\begin{figure}[t]
  \begin{scriptsize}
  \[
  \begin{array}{l}
    \text{Unit:} \quad
    \inferrule{ ~ }{\Gamma \vdash 1 \isaCat}
    \quad
    \inferrule*{ ~ }{\Gamma\pipe\alpha:C \vdash () : 1}\\\\

@@

    \text{Graph of a profunctor:} \\
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash P \isaSet}
    {\Gamma \vdash \graphProf{\alpha}{\beta} P \isaCat}
    \quad
    \inferrule
    {\defaultObCtx \vdash a_- : C_-\and
      \defaultObCtx \vdash a_+ : C_+\and
      \defaultObCtx \vdash s : P[a_-/\alpha;a_+/\beta]}
    {\defaultObCtx \vdash (a_-,a_+,s) : \graphProf{\alpha:\cat C_-}{\beta:\cat C_+} P}
    \\
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha:\cat C_-}{\beta} P}
    {\defaultObCtx \vdash \pi_- a : \cat C_-}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta:C_+} P}
    {\defaultObCtx \vdash \pi_+ a : \cat C_+}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta} P}
    {\defaultObCtx \vdash \pi_e a : P[\pi_-a/\alpha;\pi_+a/\beta]}
    \\\\
    \text{Negative Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat C \isaCat \and \cat C \isSmall}
    {\Gamma \vdash \negPresheaf {\cat C} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash a : \cat C \and
     \Gamma \pipe d^+\Xi \vdash p : \negPresheaf {\cat C}
    }
    {\Gamma \pipe \Xi \vdash \negPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \beta:\cat D \vdash \lambda \alpha:\cat C. R : \negPresheaf {\cat C}}
    \\
    \\
    \text{Positive Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat D \isaCat \and \cat D \isSmall}
    {\Gamma \vdash \posPresheaf {\cat D} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash p : \posPresheaf {\cat D}\and
     \Gamma \pipe d^+\Xi \vdash a : \cat D
    }
    {\Gamma \pipe \Xi \vdash \posPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \alpha:\cat C \vdash \lambda \beta:\cat D. R : \posPresheaf {\cat D}}
  \end{array}
  \]
  \end{scriptsize}
  \caption{Category Conectives}
  \label{fig:category-connectives}
\end{figure}

@@

Finally, we have the cartesian unit and product sets, which are
analogous to the normal unit and product of types. The most notable
point to emphasize is that in the formation rule for the product, the
two subformulae should have the same covariant and contravariant
dependence (so, as with linear logic, some constructions can
syntactically use a variable more than once and still be ``linear'').  
%% the
%% unit constructor ``uses'' all variables and the product ``uses'' all
%% variables in both cases.

\begin{figure}[t]
  % unit,
  \begin{scriptsize}
  \[
  \begin{array}{l}

@@

  \\
  \text{Cartesian unit and products:} \quad
      \inferrule{~}{\Gamma\pipe \Xi \vdash 1 \isaSet}
      \quad
      \inferrule{~}{\Gamma\pipe \Phi \vdash () :  1}
      \\ 
    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \\\\ \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}
    \quad
    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe\Phi \vdash M_i : R_i}{\Gamma \pipe\Phi \vdash (M_1,M_2) : R_1 \times R_2}
    \quad
    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}
  \end{array}
  \]
  \end{scriptsize}
  \caption{Set Connectives}
  \label{fig:set-connectives}
\end{figure}

@@

Finally, we briefly describe the connectives for the ``meta-logic'',
which extends Martin-L\"of type theory with $\Pi$/$\Sigma$ and extensional
identity types (with their standard rules)~\cite{emltt}. We use
extensional identity types so that the description of models is simpler,
but intensional identity types could be used instead.
%
The types we include are \emph{universes} for the object categorical
logic: types of small categories and locally small categories, functors,
profunctors and natural transformations.
%
The rule for the types of small categories and (large) categories are
very similar: any definable category defines an element of type $\Cats$,
and any element of that type can be reflected back into a category. The
only difference for $\smallCats$ is that the categories involved
additionally satisfy $\cat C \isSmall$. Again we elide the $\beta\eta$
principles, which state that $\quoth{-}$ and $\unquoth{-}$ are mutually
inverse.
%
Since every small category $\cat C \isSmall$ is a category $\cat C
\isaCat$, there is a definable inclusion function from $\smallCats$ to
$\Cat$ and the $\beta\eta$ properties ensure that this is a
monomorphism.

@@

\begin{figure}
  \begin{scriptsize}
  \[
  \begin{array}{l}
    \inferrule
    {~}
    {\Gamma\vdash \smallCats}
    \quad
    \inferrule
    {\Gamma \vdash \cat C \isSmall}
    {\Gamma \vdash \quoth {\cat C} : \smallCats}
    \quad
    \inferrule
    {\Gamma \vdash M : \smallCats}
    {\Gamma \vdash \unquoth M \isSmall}
    \quad
    \inferrule
    {~}
    {\Gamma\vdash \Cats}
    \quad
    \inferrule
    {\Gamma \vdash \cat C \isaCat}
    {\Gamma \vdash \quoth {\cat C} : \Cat}
    \quad
    \inferrule
    {\Gamma \vdash M : \Cat}
    {\Gamma \vdash \unquoth M \isaCat}
    \\ \\

@@

    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash t : P}
    {\Gamma \vdash \pendlambdaXdotY \alpha t : \pendallXdotY \alpha P}
    \quad
    \inferrule
    {\Gamma \vdash M : \pendallXdotY \alpha P\and
     \Gamma \pipe \beta:\cat D \vdash a : \cat C}
    {\Gamma \pipe \beta:\cat D \vdash \pendappXtoY M a : P[a/\alpha]}
  \end{array}
  \]
  \end{scriptsize}
  \caption{Type Connectives}
\end{figure}

@@

Next, we can prove that two definitions of an adjunction are equivalent:
\begin{lemma} \label{lem:adjunction}
  For $R : \varr {\cat D}{\cat C}$ and $L : \varr {\cat
    C}{\cat D}$, the following are in bijection:
  \begin{enumerate}
  \item An isomorphism of profunctors $(\punitinXfromYtoZ {\cat D} {L \alpha}{\beta}) \cong (\punitinXfromYtoZ {\cat C} \alpha {R \beta})$
  \item A unit $\eta : \pendallXdotY \alpha {\punitinXfromYtoZ {\cat C} \alpha
    {R(L\alpha)}}$ and co-unit $\varepsilon: \pendallXdotY \beta
    {\punitinXfromYtoZ {\cat D} {L(R(\beta))}{\beta}}$ satisfying triangle
    identities.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Given the forward homomorphism $\text{lr}$, we can construct $\eta =
  \pendlambdaXdotY \alpha \appunary {\text{lr}} \alpha {\punitrefl
    \alpha} {L \alpha}$ and given the unit we can reconstruct the
  forward homomorphism using
  $\text{comp}$ and $\text{fctor}$ as defined in
  Construction\ref{construction:synthetic-composition} as
  $\appbinary {\text{comp}} \alpha
  {\pendappXtoY \eta \alpha} {R(L\alpha)} {(\appunary
    {\text{fctor}(R)}{L \alpha} {f} {\beta})} {R \beta}$.
\end{proof}

\noindent We can define weighted limits, which include ordinary limits and
Kan extensions.
\begin{definition}
  For a functor $D : \varr {\cat J}{\cat C}$ and a profunctor $W :
  \harr{\cat K}{\cat J}$, the limit of $D$ weighted by $W$ is (if it
  exists) a functor $\weightedLimitDW D W : \varr{\cat K}{\cat C}$
  with an isomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {(\weightedLimitDW D W) k} \cong \homrallXYtoZ j {W k j} {(\punitinXfromYtoZ {\cat C} \alpha {D j})}$
\end{definition}
This generalizes the usual definition that a morphism into a limit is
a cone over the diagram $(\punitinXfromYtoZ {\cat C} \alpha {D j})$ to
be parameterized by a weight $W k j$.
Then we can prove the well-known theorem that right adjoints preserve (weighted) limits:
\begin{theorem}
  If $\weightedLimitDW D W$ exists and is a limit and $R : \varr{\cat
    C}{\cat C'}$ has a left adjoint $L$, then $\lambda
  \kappa. R((\weightedLimitDW D W) \kappa)$ is the limit of $\lambda
  j. R(D j)$ weighted by $W$.
\end{theorem}
\begin{proof}
\[
    \punitinXfromYtoZ {} {\gamma} {R ((\weightedLimitDW D W) \kappa)}
    \cong \punitinXfromYtoZ {}{L \gamma} {(\weightedLimitDW D W) \kappa}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{L \gamma}{D j}}
    \cong \homrallXYtoZ {j}{W k j} {\punitinXfromYtoZ {}{\gamma}{R(D j)}}\\
\]
This is a high level proof in terms of isomorphisms that may be written
in VETT. The first two steps are the instantiation of assumptions
(adjointness, weighted limits). The last step uses the fact that a
natural isomorphisms lift to natural isomorphism of homs of profunctors.
The construction of this isomorphism illustrates how naturality need not
be proved explicitly in \vett{}.  For any $\phi: \homunary \alpha {R'
  \alpha \beta} \beta {R \alpha \beta}$ and $\psi : \homunary \gamma {S
  \gamma \beta} \beta {S' \gamma \beta}$ we can construct a natural
transformation $\phi \triangleright \psi : \hombinary \gamma
{(\homrallXYtoZ \beta {R \alpha \beta}{S \gamma \beta})} {\alpha}
{R'\alpha\beta} \beta {S'\gamma\beta}$ as \\ $\lambdabinary {\gamma} {f}
{\alpha} r \beta {\appunary \psi \gamma {(\homrappXtoYatZ f {(\appunary
      \phi \alpha r \beta)} \beta)} \beta}$. Furthermore if $\phi$ and
$\psi$ have inverses, then $\phi^{-1} \triangleright \psi^{-1}$ is the
inverse of $\phi \triangleright \psi$.
\end{proof}

@@

%% MOVE TO APPENDIX examples section?

%% \subsection{Extensionality Axioms}
